---
title: Categorical Data Analysis
subtitle: template
description: |
  template
categories:
  - template
author: Kwangmin Kim
date: 03/17/2023
format: 
  html:
    page-layout: full
    code-fold: true
    toc: true
    number-sections: true
execute:
    include: false
    warning: false
    message: false
draft: False
---

<ul class="nav nav-pills" id="language-tab" role="tablist">
  <li class="nav-item" role="presentation">
    <button class="nav-link active" id="Korean-tab" data-bs-toggle="tab" data-bs-target="#Korean" type="button" role="tab" aria-controls="Korean" aria-selected="true">Korean</button>
  </li>
  <li class="nav-item" role="presentation">
    <button class="nav-link" id="English-tab" data-bs-toggle="tab" data-bs-target="#English" type="button" role="tab" aria-controls="knitr" aria-selected="false">English</button>
  </li>

<div class="tab-content" id="language-tabcontent">

<div class="tab-pane fade  show active" id="Korean" role="tabpanel" aria-labelledby="Korean-tab">

::: {#Korean .tab-pane .fade .show .active role="tabpanel" aria-labelledby="Korean-tab"}

```{r}
library(tidyverse)
```

## Goal

* Visualization Methods
    * for EDA: visualize patterns, trends, anomalies in data
    * for model diagnostic methods: visualize to assess violations of assumptions
    * for summary methods: visualize to provide an interpretable summary of data
* apply theory to practice
    * conert research questions into statistical hypotheses and models
    * look into the difference between non-parametric (ex. fisher exact test) vs parametric (ex. $\chi^2 test for independence$) vs model-based methods (ex. logistic regression)
    * for summary methods: visualize to provide an interpretable summary of data

## Definition of Categorical Data

* categorical (or frequency) data consist of a discrete set of categories, which may be ordered or unordered.
    * unordered
        * gener: {male, female, transgender}
        * marital status: {never married, married, separated, divorced, widowed}
        * party preference: {NDP, liberal, conservative, green}
        * treatment improvement: {none, some, marked}
    * ordered
        * age group: {0s,10s,20s,30s, ...}
        * number of children: {0, 1 , 2 ,3, ...}
## Structures

Categorical data appears in various forms like:

* tables
    * one way
    ```{r}
    table(sample(LETTERS[1:4], 100, TRUE))    
    ```
    * two way
    ```{r}

    ```
    * three way
    ```{r}

    ```
* matrices
    ```{r}

    ```
* array
    ```{r}

    ```
* data frames
    * case form
    ```{r}
    #| include: false
    
    data.frame(subject=1:1000,
    height=sample(170:210,1000,replace=TRUE),
    weight=sample(50:100,1000,replace=TRUE))%>%
    head(10)%>%
    knitr::kable()
    ```
    * frequency form
    ```{r}
    #| include: false
    data.frame(subject=1:1000,
    height=sample(170:210,1000,replace=TRUE),
    weight=sample(50:100,1000,replace=TRUE))%>%
    group_by(weight)%>%
    summarise(n=n())%>%head(10)%>%
    knitr::kable()
    ```

* 
:::
</div>

<div class="tab-pane fade" id="English" role="tabpanel" aria-labelledby="English-tab">

::: {#English .tab-pane .fade role="tabpanel" aria-labelledby="English-tab"}

:::


</div>
