---
title: MANOVA
subtitle: ANOVA, one-way ANOVA, Two-way ANOVA, ANCOVA, repeated measures ANOVA, MANOVA
description: | 
  The analysis of variance (ANOVA) is one of the most widely used statistical techniques. When we conduct a comparison testing of multiple groups such as A, B, and C on each with numeric data, the statistical test for a significant difference among the groups is called analysis of variance, or ANOVA.  
categories:
  - Statistics
author: Kwangmin Kim
date: 01/07/2023
draft: false
format: 
  html:
    toc: true
    number-sections: True
    code-fold: true
    page-layout: full
execute: 
  warning: false
  message: false
---


## Description

 다변량 분산분석(Multivariate Analysis of Variance, MANOVA)

- 2개 이상의 종속변수가 있을 경우 집단별  차이를 동시에 검정
- 연구의 타당성 증가

## Example

### Load Libraries and Data
```{r}
library(tidyverse)
library(faraway)
library(markdown)
library(heplots)
library(HH)
library(psych)
```

### Data Description
```{r}
str(Skulls)

```

R console에 `?Skulls`를 입력하면 다음과 같은 설명이 나온다.

**Measurements made on Egyptian skulls from five epochs.**

* The epochs correspond to the following periods of Egyptian history:
  * the early predynstic period (circa 4000 BC);
  * the late predynatic period (circa 3300 BC);
  * the 12th and 13t dynasties (circa 1850 BC);
  * the Ptolemiac peiod (circa 200 BC);
  * the Roman period(circa 150 AD).

**The question is hether the measurements change over time. Non-constant measurements of the skulls over time would indicate interbreeding with immigrant populations. Note that using polynomial contrasts for epoch essentially treats the time points as equally spaced**

즉, skulls 고대 이집트 왕조 부터 로마시대까지 이집트 지역에서 발군된 두개골의 크기를 측정한 데이터 이집트 역사를 5개의 시대로 구분하고 각 시대별로 30개씩의 두개골을 4개의 지표로 측정

이 data는 `r ncol(Skulls)`개의 변수와 `r nrow(Skulls)`개의 samples을 포함한다.

* `r names(Skulls)[1]` : 
* `r names(Skulls)[2]` : 
* `r names(Skulls)[3]` : 
* `r names(Skulls)[4]` : 
* `r names(Skulls)[5]` : 



## EDA

### Descriptive Statistics
```{r}
library(heplots)
#skulls 고대 이집트 왕조 부터 로마시대까지 이집트 지역에서 발군된 두개골의 크기를 측정한 데이터
# 이집트 역사를 5개의 시대로 구분하고 각 시대별로 30개씩의 두개골을 4개의 지표로 측정
# epoch: 이집트의 시대를 5개로 구분, 독립변수
# mb : 두개골의 폭, 종속 변수
# bh : 두개골의 높이, 종속 변수
# bl : 두개골의 길이, 종속 변수
# nh : 코의 높이, 종속 변수

library(dplyr)
sample_n(Skulls,10)

attach(Skulls)# Skulls를 작업 경로에 포함시키기
search() # 작업 경로 확인인
# 종속 변수를 결합시켜 하나의 행렬로 만들기
y<-cbind(mb,bh,bl,nh)
# 시대별 두개골  길이의 평균 보기
aggregate(y,by=list(epoch),mean) # 언뜻 보기에 차이가 있는 것 처럼 보임
# 모집단으로 일반화하기 위해 통계적 검정 시행
skulls_manova<-manova(y~epoch)
summary(skulls_manova)
# 시대별 두개골 측정값이 차이가 있는 것으로 보임

# 구체적으로 어느 두개 골 측정값에서 차이가 나는지 확인
summary.aov(skulls_manova)
# nh는 차이가 없는 것으로 보임

## 시간에 따라 두개골 측정이 다르다는 것은 이민족 유입의 혼혈 가능성이 있음

detach(Skulls)# 작업경로에서 삭제제


```



### One-Way ANOVA 



```{r}

```



## Blog Guide Map Link

* [Statistics Blog](../guide_map/index.qmd)
* [Engineering Blog](../../Engineering/guide_map/index.qmd)
* [Deep Learning Blog](../../DL/guide_map/index.qmd)
* [Machine Learning Blog](../../ML/guide_map/index.qmd)
* [Mathematics Blog](../../Mathmatics/guide_map/index.qmd)
* [Patent Blog](../../Patent/guide_map/index.qmd)
* [Validation Blog](../../Validation/guide_map/index.qmd)
